# Case Study: How Does a Bike-Share Navigate Speedy Success?

This project is about a fictional bike sharing company, _Cyclistic_, 
that aim to maximize 
the number of annual membership. 
The idea is to analyse how casual riders and annual members use 
Cyclistic bikes differently and understand how they can have the casual 
riders to buy a membership.

Cyclistic has more than 5,800 bicycles and 600 docking stations. 
It sets itself apart by offering a very large selection of bikes, 
especially they look at people with disabilities 
and riders who can’t use a standard two-wheeled bike. 
Statistically speaking, 92% of Cyclistic users use standard bikes,
moreover, only about the 30% of the total number of users use them to
commute to work each day.

The company before this analysis has  very flexible pricing plans:
single-ride passes, full-day passes, and annual memberships. 
Customers who purchase single-ride or full-day passes 
are referred to as casual riders. 
Customers who purchase annual memberships are Cyclistic members.

Since the annual membership is the most profitable, the company's director 
of marketing suggests that increasing the number of annual memberships 
is key to future growth.

The data used for the analysis are taken from
[divvy-tripdata](https://divvy-tripdata.s3.amazonaws.com/index.html) 
opensource data. The use of these data is licensed under this
[license](https://www.divvybikes.com/data-license-agreement).

## Data cleaning

Loading esseantial libraries:
```{r}
library("tidyverse")
```

To perform the required analysis I'm going to be using the data from 
the last 4 quarters; Q2, Q3 and Q4 of 2019 and Q1 of 2020.

```{r}
Q2_2019 <- 
  read_csv('C:\\Users\\er_vi\\Documents\\GitHub\\Capstone_Bike-share\\Data\\Divvy_Trips_2019_Q2.csv',
           show_col_types = FALSE)
Q3_2019 <- 
  read_csv('C:\\Users\\er_vi\\Documents\\GitHub\\Capstone_Bike-share\\Data\\Divvy_Trips_2019_Q3.csv',
           show_col_types = FALSE)
Q4_2019 <- 
  read_csv('C:\\Users\\er_vi\\Documents\\GitHub\\Capstone_Bike-share\\Data\\Divvy_Trips_2019_Q4.csv',
           show_col_types = FALSE)
Q1_2020 <- 
  read_csv('C:\\Users\\er_vi\\Documents\\GitHub\\Capstone_Bike-share\\Data\\Divvy_Trips_2020_Q1.csv',
           show_col_types = FALSE)
```

Let's start to clean data. Since we have to work with four different files 
it is first important to have consistent column namings. 
We can check how are the columns named in each file running these commands

```{r, results="hide"}
colnames(Q2_2019)
colnames(Q3_2019)
colnames(Q4_2019)
colnames(Q1_2020)
```

The files do not have all the same columns
but they share a total of nine columns. These are the columns we are going 
to merge and they need to have the same exact name.
The shared columns are: _ride_id, rideable_type, started_at, ended_at, start_station_name,_
_start_station_id, end_station_name, end_station_id_ and _member_casual_,
following the namings used in Q1_2020.

```{r, include=FALSE}
(Q4_2019 <- rename(Q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(Q3_2019 <- rename(Q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(Q2_2019 <- rename(Q2_2019
                   ,ride_id = "01 - Rental Details Rental ID"
                   ,rideable_type = "01 - Rental Details Bike ID" 
                   ,started_at = "01 - Rental Details Local Start Time"  
                   ,ended_at = "01 - Rental Details Local End Time"  
                   ,start_station_name = "03 - Rental Start Station Name" 
                   ,start_station_id = "03 - Rental Start Station ID"
                   ,end_station_name = "02 - Rental End Station Name" 
                   ,end_station_id = "02 - Rental End Station ID"
                   ,member_casual = "User Type"))

```



```{r}
str(Q1_2020)
str(Q4_2019)
str(Q3_2019)
str(Q2_2019)
```



