# Case Study: How Does a Bike-Share Navigate Speedy Success?

This project is about a fictional bike sharing company, _Cyclistic_, 
that aim to maximize 
the number of annual membership. 
The idea is to analyse how casual riders and annual members use 
Cyclistic bikes differently and understand how they can have the casual 
riders to buy a membership.

Cyclistic has more than 5,800 bicycles and 600 docking stations. 
It sets itself apart by offering a very large selection of bikes, 
especially they look at people with disabilities 
and riders who can’t use a standard two-wheeled bike. 
Statistically speaking, 92% of Cyclistic users use standard bikes,
moreover, only about the 30% of the total number of users use them to
commute to work each day.

The company before this analysis has  very flexible pricing plans:
single-ride passes, full-day passes, and annual memberships. 
Customers who purchase single-ride or full-day passes 
are referred to as casual riders. 
Customers who purchase annual memberships are Cyclistic members.

Since the annual membership is the most profitable, the company's director 
of marketing suggests that increasing the number of annual memberships 
is key to future growth.

The data used for the analysis are taken from
[divvy-tripdata](https://divvy-tripdata.s3.amazonaws.com/index.html) 
opensource data. The use of these data is licensed under this
[license](https://www.divvybikes.com/data-license-agreement).

## Data Preparation

Loading esseantial libraries:
```{r}
library("tidyverse")
library("arsenal")
library("dplyr")
library("ggplot2")
```

To perform the required analysis I'm going to be using the data from 
the last 4 quarters; Q2, Q3 and Q4 of 2019 and Q1 of 2020.

```{r}
Q2_2019 <- 
  read_csv('C:\\Users\\er_vi\\Documents\\GitHub\\Capstone_Bike-share\\Data\\Divvy_Trips_2019_Q2.csv',
           show_col_types = FALSE)
Q3_2019 <- 
  read_csv('C:\\Users\\er_vi\\Documents\\GitHub\\Capstone_Bike-share\\Data\\Divvy_Trips_2019_Q3.csv',
           show_col_types = FALSE)
Q4_2019 <- 
  read_csv('C:\\Users\\er_vi\\Documents\\GitHub\\Capstone_Bike-share\\Data\\Divvy_Trips_2019_Q4.csv',
           show_col_types = FALSE)
Q1_2020 <- 
  read_csv('C:\\Users\\er_vi\\Documents\\GitHub\\Capstone_Bike-share\\Data\\Divvy_Trips_2020_Q1.csv',
           show_col_types = FALSE)
```

Let's start to clean data. Since we have to work with four different files 
it is first important to have consistent column namings. 
We can check how are the columns named in each file running these commands

```{r}
colnames(Q2_2019)
colnames(Q3_2019)
colnames(Q4_2019)
colnames(Q1_2020)
```


These files do not have all the same column names
but they share a total of nine columns, that are: **ride_id**, **rideable_type**, **started_at**, **ended_at**, **start_station_name**,
**start_station_id**, **end_station_name**, **end_station_id** and **member_casual**,
according to the namings used in the dataset Q1_2020.

```{r, include=FALSE}
(Q4_2019 <- rename(Q4_2019,
                   ride_id = trip_id,
                   rideable_type = bikeid,
                   started_at = start_time, 
                   ended_at = end_time,
                   start_station_name = from_station_name,
                   start_station_id = from_station_id,
                   end_station_name = to_station_name,
                   end_station_id = to_station_id,
                   member_casual = usertype,
                   trip_duration = tripduration,
                   birth_year = birthyear))

(Q3_2019 <- rename(Q3_2019,
                   ride_id = trip_id,
                   rideable_type = bikeid,
                   started_at = start_time,  
                   ended_at = end_time,  
                   start_station_name = from_station_name, 
                   start_station_id = from_station_id, 
                   end_station_name = to_station_name, 
                   end_station_id = to_station_id,
                   member_casual = usertype,
                   trip_duration = tripduration,
                   birth_year = birthyear))

(Q2_2019 <- rename(Q2_2019,
                   ride_id = "01 - Rental Details Rental ID",
                   rideable_type = "01 - Rental Details Bike ID",
                   started_at = "01 - Rental Details Local Start Time",  
                   ended_at = "01 - Rental Details Local End Time",  
                   start_station_name = "03 - Rental Start Station Name", 
                   start_station_id = "03 - Rental Start Station ID",
                   end_station_name = "02 - Rental End Station Name", 
                   end_station_id = "02 - Rental End Station ID",
                   member_casual = "User Type",
                   trip_duration = "01 - Rental Details Duration In Seconds Uncapped",
                   birth_year = "05 - Member Details Member Birthday Year",
                   gender = "Member Gender"))

```

In order to use these dataframes more easily in this step I create four new 
dataframes which contain only the columns they have in common,

```{r}
Q1_2020_clean <- Q1_2020[-c(9,10,11,12)]
Q4_2019_clean <- Q4_2019[-c(5,11,12)]
Q3_2019_clean <- Q3_2019[-c(5,11,12)]
Q2_2019_clean <- Q2_2019[-c(5,11,12)]
```

and I reorganize the columns in alphabetical order for a better comparison.

```{r}
Q1_2020_clean <- Q1_2020_clean %>% select(order(colnames(Q1_2020_clean)))
Q4_2019_clean <- Q4_2019_clean %>% select(order(colnames(Q4_2019_clean)))
Q3_2019_clean <- Q3_2019_clean %>% select(order(colnames(Q3_2019_clean)))
Q2_2019_clean <- Q2_2019_clean %>% select(order(colnames(Q2_2019_clean)))
```

Now we have to check that the attributes have consistent types. 
For this we can use the str() function from the dplyr package. 

```{r}
str(Q1_2020_clean, give.attr = FALSE, give.length = FALSE, vec.len = 0)
str(Q4_2019_clean, give.attr = FALSE, give.length = FALSE, vec.len = 0)
str(Q3_2019_clean, give.attr = FALSE, give.length = FALSE, vec.len = 0)
str(Q2_2019_clean, give.attr = FALSE, give.length = FALSE, vec.len = 0)
```

From the output is clear that the type of the columns **ride_id** and **rideable_type** 
from the _Q1_2020_ is different from the other datasets. _Q1_2020_ is char 
instead of numerical. Let's convert these columns type into char.

```{r}
Q4_2019_clean <-  mutate(Q4_2019_clean, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
Q3_2019_clean <-  mutate(Q3_2019_clean, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
Q2_2019_clean <-  mutate(Q2_2019_clean, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
```

We are now ready to merge these four datasets into one.

```{r}
trips <- bind_rows(Q2_2019_clean, Q3_2019_clean, Q4_2019_clean, Q1_2020_clean)
```

## Data Cleaning


The next step is to clean our dataset **trips**. 
Let's inspect the dataset. The first thing to check is if the column **member_casual**
present any inconsistencies. Let's check what values the attributes
have,

```{r}
unique(trips["member_casual"])
```

The four values **Subscriber**, **Customer**, **member** and **casual** have to 
be mapped into only two values. 
To this end let's map **member** --> **Subscriber** 
and **casual** -->  **Customer**.

```{r}
trips <-  trips %>% 
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))
```

Next let's add a new colum in order to have a better data aggregation such 
as splitting the date of a ride into _day-month-year_.

```{r}
trips$date <- as.Date(trips$started_at)
trips$month <- format(as.Date(trips$date), "%m")
trips$day <- format(as.Date(trips$date), "%d")
trips$year <- format(as.Date(trips$date), "%Y")
trips$day_of_week <- format(as.Date(trips$date), "%A")
```

Finally, let's add other columns that indicates the duration of a trip. 
This can give insights on how customers with different memberships use the 
bike-sharing service.

```{r}
trips$ride_length <- difftime(trips$ended_at,trips$started_at) # Given in seconds
str(trips$ride_length) # Let's inspect the newly created attributes
```

The **ride_length** has a _char_ type. To make calculation with these attributes 
we have to cast them into numeric and check that the values are all bigger then
0.

```{r}
trips$ride_length <- as.numeric(as.character(trips$ride_length)) # Convert to numeric
sum(trips$ride_length <= 0)
```
Let's check the other attributes for trips with a negative duration,

```{r}
head(filter(trips,ride_length<0))
tail(filter(trips,ride_length<0))
```

Some of them have the **starting_station_name = HQ QR** that is the head quarter
where a bike has been brought for being repaired. Since these attributes 
fo not correspond to any customer it is necessary to remove them as well.
We can also check that there are other trips with a positive duration that 
started at the station **HQ QR**,

```{r}
sum(trips$start_station_name == "HQ QR") > sum(trips$ride_length <= 0)
```
Let's remove all the entries where **start_station_name** == "HQ QR" and 
**ride_length** <= 0.

```{r}
trips_clean <- trips[!(trips$start_station_name == "HQ QR" | trips$ride_length<0),]
sum(trips_clean$start_station_name == "HQ QR") # Check that the entries have been removed
sum(trips_clean$ride_length <= 0) # Check that the entries have been removed
```

This step concludes the cleaning phase of the data analysis.
Let's move onto the analysis itself.

## Analysis




```{r}
trips_clean %>%
  aggregate(ride_length ~ member_casual, FUN = mean)
```



```{r}
trips_clean$cumulative_ride_length <- 1
trips_clean %>%
  filter(ride_length < 7200 & member_casual=="casual") %>%
  arrange(ride_length) %>%
    aggregate(cumulative_ride_length ~ ride_length, FUN = sum) %>%
  ggplot(aes(x = ride_length, y = cumulative_ride_length)) +
  geom_col(position = "dodge")
```



















